---
title: "Untitled"
author: "Hilal Samut"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      # write figures to the figures folder
                      fig.path = "../figures/03_PreProcessing/")
```


# Goals 

Remove any potential contaminants and evaluate the accuracy of our sequencing run. Then, we will wite our final data objects. 

1. Load in data that we've generated: `asv_table`, `tax_table`, and metadata (which includes `metadata.csv` and `track_reads_reads`). We created these files in `analysis/02_AssignASVs.Rmd`.
2. Remove any ASVs that are chloroplasts.
3. Remove ASVs that are mitochondria. 
4. Evaluate any ASVs from the negative controls. Then, remove negative controls. 
5. Evaluate the mock community (positive control). Accuracy of sequencing.
6. Check for reverse complements. 
7. Check the sequencing depth of samples. Remove obvious sanpkes with too few reads. 

## Input

1. Metadata: `metadata.csv` and `track_read_counts.RData`.
2. ASV table:`data/01_DADA2/ASV_table.csv`
3. Taxonomy Table: `data/01_DADA2/ASV_taxonomy.csv`

## Output

1. A pre-processed phyloseq object (S4 object): `raw_preprocessed_physeq.RData`.

# Set Environment

## Load PAckages
```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")

# Load packages with pacman
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, install = FALSE)
```

## Load Data

### 1. Metadata

- What to copy?
  1. `/workdir/in_class_data/metadata.csv` -> `/workdir/netID/SalinityGradient_16S/data/`
  
  2. `/workdir/in_class_data/track_read_counts.RData` -> `/workdir/netID/SalinityGradient_16S/data/01_DADA2/`


ASV table = ASV (rows) x Sample Name (cols)
Tax = ASV (rows) x Taxonomy (cols)
Metadata = Sample Name (rows) x Info (cols)
```{load-metadata}
# load in metadata
metadata_df <-
  read_csv("data/metadata.csv") %>%
  # Fix column name 
  rename("sample_names" = "...1") %>%
  # Add sample names also as a column
  mutate(names = sample_names) 

# Inspect
head(metadata_df)
dim(metadata_df)

# include dada2 output
load("data/01_DADA2/track_read_counts.RData")

# take a look
head(track_counts_df)
dim(track_counts_df)

# Check filenames
head(track_counts_df$sample_names)

# Fix sample names in track_reads
track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names, "_"), `[`, 1)

# Intuition Check
head(track_counts_df$sample_names)

# What's different?
setdiff(track_counts_df$sample_names, metadata_df$sample_names)

# Let's do a filtering join with left_join
metadata_final_df <- 
  metadata_df %>%
  left_join(., track_counts_df, by = "sample_names") %>%
  # Sample names to the rownames to merge into phyloseq
  column_to_rownames(var = "sample_names") 

# Check
dim(metadata_final_df)
```

## 2. ASV Table

Copy data over to: `/workdir/netID/SalinityGradient_16S/data/01_DADA2/`

```{r load-asv-table}
asv_df <-
  read.delim(file = "data/01_DADA2/ASV_table.tsv", sep = "\t",
           # add Column and Row names
           header = TRUE, row.names = 1) %>%
  dplyr::select(-"CJ.V08.P")

# Inspect
asv_df[1:5, 1:5]

# Fix Column names
## Remove the X: denote at the beginning "^"
colnames(asv_df) <- sub(pattern = "^X", replacement = "", colnames(asv_df))
## Replace the . with a -: "\\." since . is a special character in regex
colnames(asv_df) <- gsub(pattern = "\\.", replacement = "-", colnames(asv_df))

# Final Inspection
head(colnames(asv_df))
asv_df[1:5, 1:5]
```

## 3. Taxonomy Table 

```{r load-tax-table}
tax_df <-
  read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t",
           header = TRUE, row.names = 1)

# Inspect
dim(tax_df)
dim(asv_df)

# Add check
stopifnot(rownames(asv_df) == rownames(tax_df))
```

# Handoff to phyloseq

```{r phyloseq-handoff}
raw_physeq <-
  phyloseq(otu_table(asv_df, taxa_are_rows = TRUE), 
         sample_data(metadata_final_df),
         tax_table(as.matrix(tax_df)))

raw_physeq

# save
save(raw_physeq, file = "data/03_PreProcessing/raw_physeq.RData")
```

# Clean up the data!

Remove chloroplasts and mitochondria

```{r rm-chloro}
noChloros_physeq <- 
  raw_physeq %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order))

# How many ASVs were chloroplasts?
numChloros_ASVs <- ntaxa(raw_physeq)- ntaxa(noChloros_physeq)
```

There were `r numChloros_ASVs` ASVs that were chloroplasts

What about mitochondria?

```{r rm-mitos}
noChlorosMitos_physeq <-
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many ASVs were mitochondria?
numMitos_ASVs <- ntaxa(noChloros_physeq)- ntaxa(noChlorosMitos_physeq)
```

There were `r numMitos_ASVs` ASVs that were mitochondria


# Evaluate and remove the control samples

```{r neg-controls}
# Make a phyloseq object with only the negative controls
neg_control_samples <- c("WaterControl", "022um-Control", 
                         "3um-Control", "DNA-Ext-Control")


neg_control_physeq <-
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% neg_control_samples) %>%
  #  rm ASVs with 0 counts
  prune_taxa(taxa_sums(.) > 0, .)

# Take a look at the ASV abundances in our negative controls
neg_control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

# What's the abundace of these neg controls ASVs in real samples?
# create vector of ASVs
neg_controlASVs_vec <-
  neg_control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()

# use vector to subset from larger dataset, creating a new phyloseq object
neg_controlASV_physeq <- 
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% neg_controlASVs_vec$ASV)

# Take a look
neg_controlASV_physeq

# Plot Abundances of neg control ASVs in samples versus controls
neg_controlASV_physeq %>%
  plot_bar(., "ASV", facet_grid = Sample_or_Control~.)

# Create a new vector with all except ASV_0017 and ASV_0050
negControlASVs_rm <- dplyr::setdiff(neg_controlASVs_vec$ASV, c("ASV_0017", "ASV_0050"))

# Make new phyloseq object with negControl ASVs removed
noChlorosMitosNegControls_physeq <-
  noChlorosMitos_physeq %>%
  # rm ASVs 
  subset_taxa(., !(ASV %in% negControlASVs_rm)) %>%
  # rm neg control samples
  subset_samples(., !(names %in% neg_control_samples))

noChlorosMitosNegControls_physeq
```





