---
title: "04A_Phylogenetics_Tree_Construction"
author: "Hilal Samut"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Goals 
The goal is to make a phylogenetic tree!

1. Load in our preprocessed phyloseq object (raw_preprocessed_phyloseq.RData)
2. Create ASV fasta file from the phyloseq object
3. Align the 16S sequences from our fasta file with MAFFT
4. Create a maximum liklihood tree using FastTree2

### Inputs
1. raw_preprocessed_phyloseq.RData


### Outputs
1. MAFT_aligned_ASV.fasta
2. ASVs_unrooted.tree



# Set Seed
```{r set-seed}
set.seed(238438)
```

# Load Packages and Phyloseq Object
```{r load-environment}
# install phytools
remotes::install_github("liamrevell/phytools")

# install ggtree
devtools::install_github("GuangchuangYu/ggtree")

pacman::p_load(phytools, ggtree, tidyverse, phyloseq, install = FALSE)

# load physeq
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")

raw_preprocessed_physeq
```

# Write ASV Fasta File

```{r create-preprocessed-fasta}
# Pull out ASV sequences and ASV names
asv_seq_df <-
  raw_preprocessed_physeq@tax_table %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# Add the ">" to make the fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

# Create a fasta object
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))


# Write a fasta file
write(asv_seq_fasta,
      file = "data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Align the 16S sequences from fasta file with MAFFT
- engine.opts='-l: gives us our original terminal base environment

To test shell directly from Rmd
- mac: command+option+enter
```{r run-maft, engine='bash', engine.opts='-l'}
# Write bash code to run MAFFT alignment

#first provide the path to MAFFT
export PATH=/programs/mafft/bin:$PATH

# change directories to provide the fasta file we made above
cd data/04_Phylogenetic_Tree
pwd

# set a seed for consistency and reproducibility
RANDOM=238428

# MAFFT already knows that we a nucleotide alignment
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta
```

# FastTree2
```{r run-fasttree, engine='bash', engine.opts='-l'}
#provide export path to FastTree
export PATH=/programs/FastTree-2.1.11:$PATH

cd data/04_Phylogenetic_Tree
pwd

# Run FastTree to generate phylogenetic tree
#parameters:
  # -nt: indicates it's a nucleotide alignment
  # -gtr: generalized time reversible substution model
  # -fastest: speed up the model, reduce model memory usage (recommended dataset that have >50,000)
  # -log: output a log file
  # input our alignment
  # specify the output tree file
  
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

# Change back to main directory
cd ../../

pwd

echo "The working directory is $PWD"
```





