---
title: "Between-Sample (Beta) Diversity of microbes along a Salinity Gradient"
author: "Hilal Samut"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../figures/06_CommunityAnalysis")
```


# Goal

1. Load in the phyloseq data with rooted tree, colors, functions.
2. Evaluate the sequencing depth and, if needed, remove samples.
3. Normalize the read counts between samples.
4. Calculate community **dis**similarities: Numbers will always mean 0 = same, 1 = dissimilar/different.
    - Sorenson
    - Bray-Curtis
    - UniFrac (weighted & unweighted)
5. Run some statistics: PERMANOVA & betadisper
6. Plot some bar plots of specific microbial groups in the data. 

## Inputs

1. Phylogenetic tree phyloseq object that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`

## Outputs

1. Calculated beta-diversity dissimilarity measures.
2. Figures: PCoA and NMDS 
3. Statistical Tests: PERMONAVA & betadisper

# Scientific Question

How does microbial composition change across a salinity gradient?

- *Null Hypothesis*: No change in microbial community with varying salinity.
- *Alternative Hypothesis*: Community compostion will change with varying salt concentrations.
   - Rare Members? Jaccard, **Sorenson**, unweighted UniFrac
   - Dominant members? **Bray-Curtis**, Weighted UniFrac
   - Evolutionary related? Unweighted or weigthed UniFrac
   

# Set up the Envireonment

## Set the seed

```{r set-seed}
set.seed(238428)
```

## Load packages

```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan,
               install = FALSE)

# Load Colors
source("code/colors.R")

# Load functions
source("code/functions.R")
```

# 1. Load in the data

```{r load-data}
# load phyloseq object 
load("data/04_Phylogenetic_Tree/phytree_preprocessed_physeqs.RData")
archaeal_rooted_physeq

# Intituion check on seq depth
min(sample_sums(archaeal_rooted_physeq))

# Create metadata_df
metadata_df <- 
  archaeal_rooted_physeq %>%
  sample_data() %>% 
  data.frame()
```

# Normalizing the Read Depth 

## 2. Explore the Raw Read Counts

```{r explore-read-counts}
# calculate read counts per sample
raw_Total_Seqs_df <- 
  archaeal_rooted_physeq %>%
  # Calculate the total number of sequences/reads
  sample_sums() %>%
  data.frame()

#Take a look
head(raw_Total_Seqs_df)

# Rename the column
colnames(raw_Total_Seqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-normalized # ASVs)
raw_TotalSeqsASVs_df <- 
  raw_Total_Seqs_df %>%
  mutate(num_ASVs = colSums(otu_table(archaeal_rooted_physeq) > 1))

#View(raw_TotalSeqsASVs_df)

# Plot histogram of seq depth 
raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0, 7500)) +
  labs(title = "Raw Seq Depth") + 
  theme_bw()

# Plot Seq Depth versus num_ASVs
raw_TotalSeqsASVs_df %>%
  ggplot(aes(x=num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 7500)) + 
  geom_smooth(method = "lm") +
  labs(title = "Seq Depth vs # ASVs")

summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))
```


# 3. Scale Read Counts

```{r scale-reads}
min(sample_sums(archaeal_rooted_physeq))

# Scale the reads
scaled_physeq <- 
  archaeal_rooted_physeq %>% 
  scale_reads(round = "matround")

# Look at it
scaled_physeq

# Confirm seq read depth of scaled_physeq

scaled_TotalSeqs_df <- 
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"

# Plot it!
scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(limits = c(0, 7500)) +
  theme_bw()

# Scaling always gives us a range!
min(scaled_TotalSeqs_df$TotalSeqs)
range(scaled_TotalSeqs_df)

41/max(scaled_TotalSeqs_df$TotalSeqs)
```

# Calculate & Visualize Community Dissimilarity 

Two different visualizations:

1. PCoA - Eigen-based analysis, linear, axes coefficients are meaningful because they represent the eigenvectors. 
2. NMDS - Non-linear method, axes are not meaningful, smush many axes of data into 2 axes

## Sorenson Dissimilarity 

### Calculate dissimilarity 

```{r soren-dis}
scaled_soren_dist <-
  phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

class(scaled_soren_dist)

## PERMANOVA: Are the centroids different?
adonis2(scaled_soren_dist ~ station, data = metadata_df)
adonis2(scaled_soren_dist ~ station + date, data = metadata_df, by = "margin")

# Betadisper
dispr_soren_station <-
  betadisper(scaled_soren_dist, metadata_df$station)
permutest(dispr_soren_station)
```

### Plot PCoA

```{r soren-pcoa}
# First, calculate PCoA with Sorenson

scaled_soren_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE)

# Take a look
str(scaled_soren_pcoa)

# Plot it!
plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_pcoa,
                color = "station",
                title = "Sorenson PCoA") +
  scale_color_manual(values = station_colors) + 
  geom_point(size = 5, alpha = 0.5, aes(color = station)) +
  theme_bw() +
  theme(legend.position = "bottom") 
```

### Plot NMDS
```{r soren-nmds}
scaled_soren_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = TRUE)

# plot it
plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_nmds,
                color = "station",
                title = "Sorenson NMDS") +
  scale_color_manual(values = station_colors) + 
  geom_point(size = 5, alpha = 0.5, aes(color = station)) +
  theme_bw() +
  theme(legend.position = "bottom") 
```





